---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
#Prepare the environment
library(knitr)
opts_chunk$set(echo = TRUE, results = 'hold')
library(data.table)
library(ggplot2)

#Unzip and load data as CSV
rdata <- read.csv('activity.csv', header = TRUE, sep = ",",            colClasses=c("numeric", "character", "numeric"))

#tidy the data
rdata$date <- as.Date(rdata$date, format = "%Y-%m-%d")
rdata$interval <- as.factor(rdata$interval)
str(rdata) # checking the data format
```

## What is mean total number of steps taken per day?

```{r}
steps_per_day <- aggregate(steps ~ date, rdata, sum)
colnames(steps_per_day) <- c("date","steps")
head(steps_per_day)

ggplot(steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "green", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 

steps_mean   <- mean(steps_per_day$steps, na.rm=TRUE)
steps_median <- median(steps_per_day$steps, na.rm=TRUE)


```
The `mean` is `r steps_mean` and the `median` is `r steps_median`.

## What is the average daily activity pattern?

```{r}
steps_per_interval <- aggregate(rdata$steps, by = list(interval = rdata$interval),FUN=mean, na.rm=TRUE)

#convert to integers
##this helps in plotting
steps_per_interval$interval <- 
        as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")
ggplot(steps_per_interval, aes(x=interval, y=steps)) +   
        geom_line(color="orange", size=1) +  
        labs(title="Average Daily Activity Pattern", x="Interval", y="Number of steps") + theme_bw()

max_interval <- steps_per_interval[which.max(steps_per_interval$steps),1]
max_number_steps <- steps_per_interval[max_interval,]
```
On average, interval value `r max_interval` contains the maximum number of steps across all the days in the dataset which is 206.2

## Imputing missing values
```{r}
missing_vals <- sum(is.na(rdata$steps))
na_fill <- function(data, pervalue) {
        na_index <- which(is.na(data$steps))
        na_replace <- unlist(lapply(na_index, FUN=function(idx){
                interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        fill_steps <- data$steps
        fill_steps[na_index] <- na_replace
        fill_steps
}

rdata_fill <- data.frame(  
        steps = na_fill(rdata, steps_per_interval),  
        date = rdata$date,  
        interval = rdata$interval)
str(rdata_fill)
sum(is.na(rdata_fill$steps)) # checking missing value

fill_steps_per_day <- aggregate(steps ~ date, rdata_fill, sum)
colnames(fill_steps_per_day) <- c("date","steps")

##plotting the histogram
ggplot(fill_steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "blue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 



steps_mean_fill   <- mean(fill_steps_per_day$steps, na.rm=TRUE)
steps_median_fill <- median(fill_steps_per_day$steps, na.rm=TRUE)
```

The imputed data mean is `r steps_mean_fill`
The imputed data median is `r steps_median_fill`

## Are there differences in activity patterns between weekdays and weekends

```{r}
weekdays_steps <- function(data) {
    weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
                          FUN=mean, na.rm=T)
    # convert to integers for plotting
    weekdays_steps$interval <- 
            as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
    colnames(weekdays_steps) <- c("interval", "steps")
    weekdays_steps
}

data_by_weekdays <- function(data) {
    data$weekday <- 
            as.factor(weekdays(data$date)) # weekdays
    weekend_data <- subset(data, weekday %in% c("Saturday","Sunday"))
    weekday_data <- subset(data, !weekday %in% c("Saturday","Sunday"))

    weekend_steps <- weekdays_steps(weekend_data)
    weekday_steps <- weekdays_steps(weekday_data)

    weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
    weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))

    data_by_weekdays <- rbind(weekend_steps, weekday_steps)
    data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
    data_by_weekdays
}

data_weekdays <- data_by_weekdays(rdata_fill)

ggplot(data_weekdays, aes(x=interval, y=steps)) + 
        geom_line(color="violet") + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of steps") +
        theme_bw()
```

